---
title: "RNAseq Sandbox"
output: html_notebook
---

### RNA-seq sandbox

Welcome to the "sandbox" of the RNA-seq course. This is a notebook containing all of the course code with none of the explanation contained in the instructional materials. Use this as a space to manipulate variables, try different thresholds/values in the functions, etc. Please note that any chunks labeled "* boilerplate *" are best left untouched since manipulating values within this code will not yeild interesting results.

> ### >>> TRY:

This comment is a flag for experimentation. For example:

```{r this chunk does nothing}
### >>> TRY: setting x = 10
x <- 4
rm(x)
```

Happy exploring!

### Setup
```{r message=FALSE, echo=FALSE, collapse=TRUE}
## Defining install_load function - adapted from: https://gitlab.com/iembry/install.load
install_load <- function(package1, ...) {   
   # convert arguments to vector
   packages <- c(package1, ...)
   # start loop to determine if each package is installed
   for(package in packages) {
       # if package is installed, just load it
       if(package %in% (BiocManager::available()))
          do.call('library', list(package))
       # if package is not installed locally, download, then load
       else {
          BiocManager::install(package)
          do.call("library", list(package))
       }
   }    
}
install_load("GEOquery","org.Mm.eg.db","GenomeInfoDbData","TxDb.Mmusculus.UCSC.mm10.knownGene","Mus.musculus","BiocStyle","limma","edgeR","RColorBrewer","gplots","gage","pathview")
```

Set the directory where you want the data to be stored (this assumes that you have a folder called _RNAseq_ on your desktop, with a subfolder called _Data_). 
```{r setup *boilerplate*}
## if you ARE using an R notebook (if not the comment out this line):
knitr::opts_knit$set(root.dir = "~/Desktop/RNAseq/Data/") 

## if you ARE NOT using an R notebook (un-comment this line):
#setwd("~/Desktop/RNAseq/Data/") 
```

# >>> Downloading and prepping the data

### Download data from GEO
```{r download data *boilerplate*}
gse <- getGEOSuppFiles(GEO = "GSE63310", makeDirectory = TRUE)

untar("./GSE63310/GSE63310_RAW.tar")

files <- c("GSM1545535_10_6_5_11.txt", "GSM1545536_9_6_5_11.txt", "GSM1545538_purep53.txt", "GSM1545539_JMS8-2.txt", "GSM1545540_JMS8-3.txt", "GSM1545541_JMS8-4.txt", "GSM1545542_JMS8-5.txt", "GSM1545544_JMS9-P7c.txt", "GSM1545545_JMS9-P8c.txt")

for(i in paste(files, ".gz", sep="")) {
    gunzip(i, overwrite=TRUE)
}

f1 <- read.delim(files[1])

dge <- readDGE(files, columns=c(1,3))
dge
```

### Anotate data
```{r annotate *boilerplate*}
dge$samples$group <- as.factor(c("LP", "ML", "Basal", "Basal", "ML", "LP", "Basal", "ML", "LP"))

dge_ids <- rownames(dge)

gene_ids <- select(org.Mm.eg.db, 
    keys=dge_ids, 
    keytype="ENTREZID",
    columns=c("SYMBOL"))

dge$gene_ids <- gene_ids
```

# >>> Processing and analyzing the data

### Converting raw counts to CPM
```{r convert to cpm}
f1 <- read.delim(files[1])

cpm <- cpm(dge)

### >>> TRY: use the help function to try different 
# ?log2_cpm
log2_cpm <- cpm(dge, log=TRUE)

# Good place to stop and check that everything looks right
#head(dge$counts)
#head(cpm)
#head(log2_cpm)
#View(dge)
```

### Filter out low-read genes
```{r filtering genes with low intensity *edit to affect analysis*}
### >>> TRY: change filtering thresholds
dge_sub <- dge[rowSums(cpm>1)>=3, ]
```

Confused? Break this down!
```{r check thresholds *edit to explore filtering thresholds*}
### >>> TRY: change filtering thresholds
head(cpm)
head(rowSums(cpm>1))
head(rowSums(cpm>1)>=3)
```

**Think:** What are some ways to decide on the filtering criteria? How do you do it? What are the strengths and weaknesses of these options:   

* Intensity-based (e.g. decide a-priori: if genes are not expressed in x% of samples, remove).  
* Statistically-based (e.g. calculate coefficient of variation CV or adjusted p-values for each gene across samples, decide a-priori to remove genes outside of a certain range). 
* Biologically-based (e.g. look at genes of interest).

```{r}
nsamples <- ncol(dge_sub)
col <- brewer.pal(nsamples, "Paired")
col
### >>> TRY: use ?cpm to modify this function call
log2cpm_raw <- cpm(dge, log=TRUE)
log2cpm_sub <- cpm(dge_sub, log=TRUE)
```

### Plot raw vs filtered data
```{r plot raw vs unfiltered}
par(mfrow=c(1,2))
plot(density(log2cpm_raw[,1]),
     col=col[1], 
     lwd=2, 
     ylim=c(0,0.21), 
     las=2, 
     main="", 
     xlab="")
title(main="A. Raw data", xlab="Log2(CPM)")
abline(v=0, lty=3)

for (i in 2:nsamples){
    den <- density(log2cpm_raw[,i])
    lines(den$x, den$y, col=col[i], lwd=2)
}

plot(density(log2cpm_sub[,1]), 
    col=col[1], 
    lwd=2, 
    ylim=c(0,0.21), 
    las=2, 
    main="", 
    xlab="")
title(main="B. Filtered data", xlab="Log2(CPM)")
abline(v=0, lty=3)

for (i in 2:nsamples){
    den <- density(log2cpm_sub[,i])
    lines(den$x, den$y, col=col[i], lwd=2)
}
```
**Think:** why is there that peak on the left side of the raw data? Recall the `cpm` function in chunk #6. Use `?cpm` - check input values and how they are handled by default. You'll notice that if you manipulated the filtering steps above, these plots may have changed from what they showed in class. How to these differ?

### Normalize data
```{r introducing variance to demo normalization}
# this normalizes the data
dge_sub_norm <- calcNormFactors(dge_sub, method = "TMM")

# this code skews data to make the effects of normalization more noticable - would not be done in a real analysis
dge_sub$counts[,1] <- dge_sub$counts[,1]*0.05
dge_sub$counts[,2] <- dge_sub$counts[,2]*5

# save datasets
log2cpm_unorm <- cpm(dge_sub, log=TRUE)
log2cpm_norm <- cpm(dge_sub_norm, log=TRUE)
```

```{r plot normalized vs non-normaliezd data}
par(mfrow=c(1,2))
boxplot(log2cpm_unorm, 
    las=2, 
    col=col, 
    main="")
title(main="A. Unnormalized data",ylab="Log2(CPM)")

boxplot(log2cpm_norm, 
    las=2, 
    col=col, 
    main="")
title(main="B. Normalized data",ylab="Log2(CPM)")
```

### Unsupervised clustering

```{r clustering}
### assign colors to groups
grps <- dge_sub$samples$group
col.grp <- grps
levels(col.grp) <-  brewer.pal(nlevels(col.grp), "Set1")
col.grp <- as.character(col.grp)

### plotting via `plotMDS`
par(mfrow=c(1,1))
plotMDS(log2cpm_norm, 
    labels=grps, 
    col=col.grp,
    xlab="Log2(Fold-Change)",
    ylab="Log2(Fold-Change)")
title(main="Sample groups")
```

# >>> Analysis and visualization

### Differential expression analysis
```{r differential expression}
design <- model.matrix(~0 + grps)

colnames(design) <- gsub("grps", "", colnames(design))

### make contrast matrix
contrast <- makeContrasts(
   BasalvsLP = Basal - LP, 
   BasalvsML = Basal - ML, 
   LPvsML = LP - ML, 
   levels = colnames(design))

### VOOM
dge_voom <- voom(dge_sub_norm, design)

# fit linear model
dge_fit <- lmFit(dge_voom, design)
# compare groups using contrast matrix
grp_fit <- contrasts.fit(dge_fit, contrasts=contrast)

head(dge_fit$coefficients)
head(grp_fit$coefficients)

efit <- eBayes(grp_fit)
efit

### plot mean-variance trend
plotSA(efit, main="Mean−variance trend")
```
**Think:** Would you exclude any genes based on this plot? 

```{r find significantly variable genes}
summary(decideTests(efit))
```

```{r top 10 genes by F statistics and pvals}

## look at top 10 overall by F statistics/p-values
topTableF(efit, number=10) 

names(efit)
efit$contrasts

## Basal vs. LP (coef=1)
topTable(efit, coef=1, number=10, sort.by="p")

## Basal vs. ML (coef=2)
topTable(efit, coef=2, number=10, sort.by="p")

## LP vs. ML (coef=3)
topTable(efit, coef=3, number=10, sort.by="p")

### save results:
#write.fit(efit, file="DEresults.txt", adjust="BH", method="separate")
```

### Visualize data

```{r heatmaps}
var_genes <- apply(log2cpm_norm, 1, var)

# select top n number of variable genes
select_var <- names(sort(var_genes, decreasing=TRUE))[1:500]

highly_variable_lcpm <- log2cpm_norm[select_var,]

## Get some nicer colors
## to display brewer colors (use diverging colors in this instance): display.brewer.all()
mypalette <- brewer.pal(11,"RdBu") 
morecols <- colorRampPalette(mypalette)
# Set up color vector for celltype variable
col.cell <- c("honeydew4", "lightblue3", "lightcyan2")[c(dge$samples$group)]

# png(filename="heatmap.png")
heatmapp <- heatmap.2(highly_variable_lcpm, col=rev(morecols(50)),
          trace = "none",
          main = "Top 500 most variable genes - ML v. LP v. Basal",
          keysize = 1.75,
          key.title = "",
          lwid = c(0.09, 0.4), # helps format the legend - arbitrary
          ColSideColors = col.cell,scale="row",
          srtCol = 25,
          labRow = NA,
          margins = c(8,8),
          )
# dev.off()

heatmapp
# view dendograms, heatmap

plot(heatmapp$colDendrogram)
plot(heatmapp$rowDendrogram)
```

```{r heatmaps exploring the dendogram}
top_500_clustered <- as.hclust(heatmapp$rowDendrogram)

# define the clusters (essentially picking where to "cut"" the dendogram and select clusters)
# cutree() returns a vector of cluster membership in the order of the original data rows
mycl <- cutree(top_500_clustered, h=max(40))

# examine it
#mycl

# examine the head and tail of cluster membership by it's order in the heatmap
head(mycl[top_500_clustered$order], n=10)
tail(mycl[top_500_clustered$order], n=10)

# you could either grab a cluster
#cluster1 <- highly_variable_lcpm[mycl == 1,]

# or simply add the cluster ID to your data. This is what we'll do.
cldat <- cbind(highly_variable_lcpm, clusterID=mycl)

# examine the data with cluster ids attached, and ordered like the heat map
top_diff_genes <- as.data.frame(cldat[top_500_clustered$order,])

# check is gene IDs table is still in memory:
head(gene_ids)

#use gene IDs to rename genes in hr...
top_diff_genes <- tibble::rownames_to_column(top_diff_genes, "ENTREZID")
top_diff_genes <- merge(gene_ids, top_diff_genes, by="ENTREZID")
head(top_diff_genes, n=10)

## EXAMPLE - view genes only in cluster 1
### >>> TRY: edit cutree() call to change location where dendogram is cut
mycl <- cutree(top_500_clustered, h=max(40))
head(mycl[top_500_clustered$order], n=10)
tail(mycl[top_500_clustered$order], n=10)

# Now we'll grab a cluster
cluster1 <- highly_variable_lcpm[mycl == 1,]
# as opposed to adding the cluster ID to the data, as we did previously.
#cldat <- cbind(highly_variable_lcpm, clusterID=mycl)

top_diff_genes <- as.data.frame(cluster1)

#use gene IDs to rename genes in hr...
top_diff_genes <- tibble::rownames_to_column(top_diff_genes, "ENTREZID")
top_diff_genes <- merge(gene_ids, top_diff_genes, by="ENTREZID")
head(top_diff_genes, n=10)

# And if we are interested in further dissecting genes by cluster, we may adjust where that threshold is as follows:

mycl <- cutree(top_500_clustered,
               h=max(20) # here is the line to edit in this case - sets the height at which to "cut" the dendogram
               )

head(mycl[top_500_clustered$order], n=10)
tail(mycl[top_500_clustered$order], n=10)

# This time we're adding the cluster assignment to the data to observe what this threshold adjustment actually accomplished
cldat <- cbind(highly_variable_lcpm, clusterID=mycl)

top_diff_genes <- as.data.frame(cldat[top_500_clustered$order,])
top_diff_genes <- tibble::rownames_to_column(top_diff_genes, "ENTREZID")
top_diff_genes <- merge(gene_ids, top_diff_genes, by="ENTREZID", sort = F)
head(top_diff_genes)
```

### Volcano plots

```{r reshape data}
### Rehashing how to format data for volano plots (from notebooks 6, 7, 8)
# start with our groups
grps # this variable contains our groups of samples (ML, LP, Basal) - created in notebook #7 (unsupervised clustering)
# creating design matrix (dummy variable creation for each sample and the group it belongs to)
design <- model.matrix(~0 + grps)
# relabeling columns
colnames(design) <- gsub("grps", "", colnames(design))
#creating contrast matrix
contrast <- makeContrasts(
   BasalvsLP = Basal - LP, 
   BasalvsML = Basal - ML, 
   LPvsML = LP - ML, 
   levels = colnames(design))
# Now we use `voom()` from `limma` to combine log2(CPM) values from the filtered, normalized data with the design matrix
dge_voom <- voom(dge_sub_norm, design)
# Using limma's lmfit() function - this is the linear model fit step
dge_fit <- lmFit(dge_voom, design)
# Using contrast matrix to compare the groups
grp_fit <- contrasts.fit(dge_fit, contrasts=contrast)
# Empirical Bayes method (limma) to use information across all genes - this should make the residual variances independent of the expression levels.
efit <- eBayes(grp_fit)


# convert gene IDs to Gene Symbols for easier interpretation
efit_genes <- names(efit$Amean)
efit_genes <- as.data.frame.AsIs(efit_genes)
colnames(efit_genes) <- "ENTREZID"
efit_genes <- merge(gene_ids, efit_genes, by="ENTREZID", sort = F)
names(efit$Amean) <- efit_genes$SYMBOL
```

```{r plot volcanoplot}
# basal vs lp
volcanoplot(efit,
            coef = 1,
            style = "p-value",
            highlight = 10,
            names = names(efit$Amean),
            hl.col="blue",
            main = "Basal vs LP",
            xlab = "Log2 Fold Change",
            ylab = NULL,
            pch=20,
            cex=0.35)

# basal vs ml
volcanoplot(efit, coef = 2, style = "B-statistic", highlight = 10, names = names(efit$Amean), 
            hl.col="blue",
            main = "Basal vs ML", xlab = "Log2 Fold Change", ylab = NULL, pch=20, cex=0.35)

# lp vs ml
volcanoplot(efit,coef = 3, style = "p-value", highlight = 10, names = names(efit$Amean), 
            hl.col="blue", 
            main = "LP vs ML", xlab = "Log2 Fold Change", ylab = NULL, pch=20, cex=0.35)
```

```{r haetmap followup analysis}
# top hits:
BasalvLP_top <- topTable(efit, coef = 1, number = 10, sort.by = 'P')
#BasalvLP_top

# top hits:
BasalvML_top <- topTable(efit, coef = 2, number = 10, sort.by = 'P')
#BasalvML_top

# top hits:
ML_v_LP_top <- topTable(efit, coef = 3, number = 10, sort.by = 'P')
#ML_v_LP_top

# can convert gene IDs to Gene Symbols for easier interpretation
ML_v_LP_top <- tibble::rownames_to_column(ML_v_LP_top, "ENTREZID")
ML_v_LP_top <- merge(gene_ids, ML_v_LP_top, by="ENTREZID", sort = F)
ML_v_LP_top


```

### Geneset enrichment analysis

```{r gsea}
fold_lpml <- efit$coefficients[, "LPvsML"]

kegg_mouse <- kegg.gsets(species="mmu", id.type="kegg")

kegg_all = gage(exprs=fold_lpml, gsets=kegg_mouse$kg.sets, same.dir=TRUE)

lapply(kegg_all, head)

kegg_mouse$kg.sets[1] ## this is the first metabolic pathway (index=1)

kegg_met <- kegg_mouse$kg.sets[kegg_mouse$met.idx] ## all metabolic pathways

kegg_met = gage(exprs=fold_lpml, gsets=kegg_met, same.dir=TRUE)

## extract KEGG ID and pathway name (first row name of the kegg_met$less data)
op <- rownames(kegg_met$less)[1]
op

## extract the KEGG ID (this code removes everything after the first space from op)
op_kegg <- sub(" .*$", "", op)
op_kegg

# plot pathway
pathview(gene.data=fold_lpml, pathway.id=op_kegg, species="mmu")
```

