---
output: html_notebook
---

####*Objective 3. Calculate expression as read counts/million (CPM) and log2(CPM)*

<br/>
Packages used in this notebook: `edgeR`. First verify that it is installed and loaded.
```{r message=FALSE, echo=FALSE}
## Defining install_load function - adapted from: https://gitlab.com/iembry/install.load
install_load <- function(package1, ...) {   
   # convert arguments to vector
   packages <- c(package1, ...)
   # start loop to determine if each package is installed
   for(package in packages) {
       # if package is installed, just load it
       if(package %in% (BiocManager::available()))
          do.call('library', list(package))
       # if package is not installed locally, download, then load
       else {
          BiocManager::install(package)
          do.call("library", list(package))
       }
   }    
}
install_load("edgeR")
```

<br/>
Set the directory where you have the data to be stored (this assumes that you have it in a folder called _RNAseq_ on your desktop, with a subfolder called _Data_). 
```{r setup}
## if you ARE using an R notebook (if not the comment out this line):
knitr::opts_knit$set(root.dir = "~/Desktop/RNAseq/Data/") 

## if you ARE NOT using an R notebook (un-comment this line):
#setwd("~/Desktop/RNAseq/Data/") 
```

<br/>
We will now convert the raw counts into counts per million (CPM). We will also be calculating log2 counts/million, log2(CPM), which will be what we primarly analyze. Note that sometimes other transformations of counts are used such as RPKM and FPKM, which account for gene length differences (shown in individual files). In this experiment, we only care about expression differences between the sample groups, and are assuming that expressed genes do not have different gene lengths overall (so we'll stick with CPM). 

<br/>
Now let's take another look at part of the first raw file (which is tab-separated).
```{r}
f1 <- read.delim(files[1])
dim(f1)
head(f1)
```

<br/>
The `cpm()` function from the `edgeR` package calculates CPM values using the DGEList object that we created in Objective 2. It can also calculate log2(CPM) values. 
```{r}
?cpm
cpm <- cpm(dge)
log2_cpm <- cpm(dge, log=TRUE)
```

<br/> 
Take a look at raw counts, raw CPM, and log2(CPM) values to make sure all looks good.
```{r}
head(dge$counts)
head(cpm)
head(log2_cpm)
View(dge)
```
Note: when a raw count is zero, the `cpm()` function will add the average of the counts to avoid talking the log of zero, which is not a number. We'll be filtering these out anyway.

