---
output:
  html_notebook:
    df_print: paged
---

####*Objective 5. Normalize the filtered expression data and plot*

<br/>
Packages used in this notebook: `edgeR`, `RColorBrewer`. First verify that they are installed and loaded.
```{r message=FALSE, echo=FALSE}
## Defining install_load function - adapted from: https://gitlab.com/iembry/install.load
install_load <- function(package1, ...) {   
   # convert arguments to vector
   packages <- c(package1, ...)
   # start loop to determine if each package is installed
   for(package in packages) {
       # if package is installed, just load it
       if(package %in% (BiocManager::available()))
          do.call('library', list(package))
       # if package is not installed locally, download, then load
       else {
          BiocManager::install(package)
          do.call("library", list(package))
       }
   }    
}
install_load("edgeR","RColorBrewer")
```

<br/>
Now we should normalize the filtered expression data in case there were any issues with our samples. These values can go in the `norm.factors` column of the samples part of our filtered DGEList object `dge_sub`. 
```{r}
dge_sub$samples
```

<br/>
There are a number of different packages that will do this (some may work better than others!), but we'll use the `calcNormFactors()` from the `edgeR` package.
```{r}
?calcNormFactors
dge_sub_norm <- calcNormFactors(dge_sub, method = "TMM")
dge_sub_norm$samples
```

<br/>
Let's see what happens when we mess up the scale on the first 2 columns of the raw filtered counts (reduce counts by 5% for first sample, increase by 500% for second). We will then re-calculate the log2(CPM) values after doing this. This will better allow us to see the effects of the normalization in the plots below.

```{r}
dge_sub$counts[,1] <- dge_sub$counts[,1]*0.05
dge_sub$counts[,2] <- dge_sub$counts[,2]*5

log2cpm_unorm <- cpm(dge_sub, log=TRUE)
log2cpm_norm <- cpm(dge_sub_norm, log=TRUE)
```

<br/>
Now we can make side-by-side box plots to show the differences before and after normalization (especially for the two samples that we messed up intentionally).   
```{r}
par(mfrow=c(1,2))
boxplot(log2cpm_unorm, 
    las=2, 
    col=col, 
    main="")
title(main="A. Unnormalized data",ylab="Log2(CPM)")

boxplot(log2cpm_norm, 
    las=2, 
    col=col, 
    main="")
title(main="B. Normalized data",ylab="Log2(CPM)")
```


