---
output: html_notebook
---

####*Objective 7. Perform differential expression analysis*

<br/>
Packages used in this notebook: `limma`, `RColorBrewer`. First verify that they are installed and loaded.
```{r message=FALSE, echo=FALSE}
## Defining install_load function - adapted from: https://gitlab.com/iembry/install.load
install_load <- function(package1, ...) {   
   # convert arguments to vector
   packages <- c(package1, ...)
   # start loop to determine if each package is installed
   for(package in packages) {
       # if package is installed, just load it
       if(package %in% (BiocManager::available()))
          do.call('library', list(package))
       # if package is not installed locally, download, then load
       else {
          BiocManager::install(package)
          do.call("library", list(package))
       }
   }    
}
install_load("limma","RColorBrewer")
```

<br/>
Set the directory where you want the results to be stored (this assumes that you have a folder called _RNAseq_ on your desktop, with a subfolder called _Results_). 
```{r setup}
## if you ARE using an R notebook (if not the comment out this line):
knitr::opts_knit$set(root.dir = "~/Desktop/RNAseq/Results/") 

## if you ARE NOT using an R notebook (un-comment this line):
#setwd("~/Desktop/RNAseq/Results/") 
```

<br/>
Since clustering suggested there may be differences in sample groups, here we will be formally testing whether genes are differentially expressed between the groups.  
First we need to do a little set-up which involves creating design and contrast matrices.

<br/>
Let's make a design matrix with groups to be compared (this makes dummy variables for each sample indicating which group they belong to). We can do this with the `model.matrix()` function from the `stats` package.
```{r}
?model.matrix
grps
design <- model.matrix(~0 + grps)
design
```

<br/>
Now we'll just relable the columns.
```{r}
colnames(design) <- gsub("grps", "", colnames(design))
design
```

<br/>
We also need a contrast matrix to indicate which groups to compare to each other (reference group has a value of 1, comparison group has a value of -1). The `makeContrasts()` function is from the `limma` package.

```{r}
?makeContrasts

contrast <- makeContrasts(
   BasalvsLP = Basal - LP, 
   BasalvsML = Basal - ML, 
   LPvsML = LP - ML, 
   levels = colnames(design))
contrast
```

<br/>
Now we use `voom()` from `limma` to combine log2(CPM) values from the filtered, normalized data with the design matrix.
```{r}
?voom

dge_voom <- voom(dge_sub_norm, design)
dge_voom
```

<br/>
Now we can put this into a linear model where the groups are compared separately for each gene using `lmFit()` from `limma`.
```{r}
?lmFit
dge_fit <- lmFit(dge_voom, design)
```

<br/>
Next we'll use the contrast matrix created above so we can compare the groups to each other (this is what we care about the most).
```{r}
?contrasts.fit
grp_fit <- contrasts.fit(dge_fit, contrasts=contrast)
```


<br/>
Notice that these grp_fit coefficients are equal to the differences between the coefficients from the dge_fit model above.
```{r}
head(dge_fit$coefficients)
head(grp_fit$coefficients)
```

<br/>
Finally, we use an empirical Bayes method to use information across all genes- this should make the residual variances independent of the expression levels. The `eBayes()` function is also from `limma`.
```{r}
?eBayes
efit <- eBayes(grp_fit)
efit
```

<br/>
Just to make sure the eBayes method worked, we can plot the average expression vs. the residual variance (there should be no correlation).
```{r}
?plotSA
plotSA(efit, main="Mean−variance trend")
```
**Question:** Would you exclude any genes based on this plot?  

<br/>
Now we can find out how many genes are significantly different between the groups.
```{r}
?decideTests
summary(decideTests(efit))
```

<br/>
And we can find the most significant genes (e.g. top 10). The F statistic reflects the magnitude of the overall group differences, and the adjusted p-value corrects for multiple comparisons. It is possible to look at overall statistics for the three contrasts, but we will take a look at the top 10 genes for each contrast one at a time (i.e. Basal vs. LP, Basal vs. ML, LP vs. ML) by using the coef argument.  
```{r}
?topTable

## look at top 10 overall by F statistics/p-values
topTableF(efit, number=10) 

names(efit)
efit$contrasts

## Basal vs. LP (coef=1)
topTable(efit, coef=1, number=10, sort.by="p")

## Basal vs. ML (coef=2)
topTable(efit, coef=2, number=10, sort.by="p")

## LP vs. ML (coef=3)
topTable(efit, coef=3, number=10, sort.by="p")
```

<br/>
If you would like to save the results from the model above, including coefficients, test statistics and p-values, you can use the `write.fit()` function from the `limma` package. In this function, you need to specify if you want to adjust p-values for multiple comparisons (you should, and we will use "BH" here for Benjamini-Hochberg, to match what we did above with topTable, and with what we will do next with visualizations and GSEA). We will also indicate that we want to adjust p-values for each contrast.

```{r}
write.fit(efit, file="DEresults.txt", adjust="BH", method="separate")
```

